### 2005-02-03 doc/paper/skysubtraction.tex
latest update

A global sky model is fit using all available fiber spectra assigned as 
"SKY" by target selection.  These fibers are set on blank fields, and are 
used to estimate the sky level as a function of wavelength and CCD position 
in individual exposures.   Each plate is nominally assigned 32 of its 640 
fibers as "SKY", with the focal plane positions chosen to allow half of the 
sky fibers to reach each spectrograph: 1 fiber per bundle of 20 is assigned to 
"SKY".

The sky subtraction is done after the data have been extracted from 
two-dimensional counts per CCD pixel to one-dimensional "corrected" electron 
counts as a function of vacuum heliocentric wavelengths and fiber number.
The flat-field "correction" removes differences due to surface brightness 
changes due to varying spectral dispersion in different fibers.

A single subroutine is called three consecutive times to arrive at the final 
model description for the sky background.  The first call uses all available sky
fibers that were not masked as an entire fiber (e.g. by the fibermask bits).
If only one good skyfiber is found, the routine logs an "ABORT" and returns.
A median count level is summarized for the selected sky spectra, and the 5% 
and 95% levels are reported for each science exposure (on each camera).
A single one-dimensional sky model is created by a b-spline fit to the airmass 
corrected sky spectra (See appendix b-spline for a continued description).
The airmass associated with the exposure mid-point (in time) is divided out 
of each sky spectrum (and errors scaled accordingly).  The differential airmass
across a single SDSS plate is the largest source of residuals after the dominant
wavelength residuals.  Individual pixels are masked if they were divided by a 
low flat field vector or a bad pixel was detected nearby (clarify?).
The number of discarded pixels before sky modeling is logged.

The sky

>>>>

Sky subtraction is performed on a frame by frame basis using the 
"flattened" wavelength calibrated extractions.  The goal of sky subtraction
is to estimate the expected background counts contributing to each of 
the native 320x2048 pixels in each extracted frame.  A "supersky" model 
at a fiducial airmass of 1 is constructed by performing iterative fits to
the extracted spectra associated with blank sky positions (so-called
"sky" fibers).   The following steps are taken in 3 sequential iterations
in order to produce the sky-subtracted spectra:

\begin{enumerate}

\item{Identify the "sky" fibers which have no bits flagged 
in the fiber mask array.}

\item{Divide each sky spectrum by the airmass of the fiber position appropriate
for the midpoint of the observation to produce respresentative sky spectra for 
an airmass of 1.  The inverse variance is scaled accordingly.}

\item{Group and sort the sky spectra as a function of wavelength.  This step
transforms a 2-dimensional set of N_sky x 2048 pixels into a sorted 1-dimensional array.  Each pixel maintains its associated central wavelength and 
inverse variance. }

\item{Perform a single cubic b-spline fit as a function of wavelength.  
The number of breakpoints (number of free parameters is 4 greater) is set 
at 2048 on the red side and 1536 on the blue side.  The breakpoint spacing 
is set automatically to maintain approximately constant S/N between breakpoint 
positions. The b-spline fit itself is iterative, with upper and lower rejection thresholds set to mask bad or deviant pixels. }

\item{The b-spline fit is evaluated at all 320x2048 wavelengths representing all
spectral pixels in the frame.  The airmass correction is applied for each individual fiber, and the rescaled super-sky is subtracted from each spectrum.}

\item{Estimate the significance of the residuals in the sky fibers used for the
fit in wavelength bins separated by breakpoints determined above.  for each bin
where inverse the 67th percentile of $\chi^2$ is greater than 1, we rescale 
all the inverse variance values down by the 67th percentile value of $\chi^$.
The rescaling is actually done with a smooth interpolation as a function of 
redshift with a constraint that the inverse variances cannot increase. }

%The b-spline routines from the numerical fortran library: {\bf slatec}
%were ported to IDL to be easily incorporated into the pipeline.

The sequence followed in the determination of the sky subtracted spectrum
and re-scaled errors is called 3 times successively.  After the 1st call,
any sky fibers which have a median \chi^2 per pixel greater than 2 is 
masked as a 'BADSKYFIBER' and the corresponding bit is set in the fibermask
array.  If any fibers are masked, the sky subtraction is iterated a 2nd time.
The sky subtraction routine is invoked for one final iteration, but with extra 
freedom allowed in the fit.  The same freedom is allowed in wavelength, by
keeping the same number of breakpoints, but a quadratic polynomial fit is
allowed as a function of wavelength.  This triples the number of free 
parameters in the linear least-squares fit, but the values are still 
overconstrained due to the ample number of sky fibers.  The extra freedom
in the model can accomodate variation in the PSF of sky features as a function 
of position on the CCD.

The sky subtraction method which utilizes the b-spline model takes advantage 
of the inherent over-sampling in wavelength due to the large number of 
sky fibers assigned per plate, without the requirement of resampling the 
pixel value to a fixed linear wavelength scale.  The least squares fit
is carried out for native spectral pixels where the errors are very nearly 
uncorrelated and the fit is evaluated for all the native pixels (320x2048)  
at the central wavelengths of the single spectral frames.

The accuracy of the SDSS spectral sky subtraction has been tested in recent
scientific analyses using large numbers of background-limited spectra.   
The distribution of flux residuals between spectral observations 
on separate nights has been shown to be 6\%-9\% larger in width than expected
from the reported errors (McDonald et al. 2004, Burgess \& Burles 2005).
For combined observations of faint galaxies, the distribution of residuals 
about the best model fit is well described by a gaussian with exponential
tails (Bolton et al. 2004), where the magnitude of tails is greatly enhanced in
the vicinity of strong sky emission lines.
